---
title: "R Notebook"
output:
  html_notebook: default
  pdf_document: default
---

##### Dataset: Online Retail.xlsx
##### Source: [Coursera] (https://www.coursera.org/projects/perform-exploratory-data-analysis-on-retail-data-with-python)
##### Objectives: 
1. Assess sales trends over time (daily, monthly, yearly, day of week, seasons)
2. Assess sales trends by country (top-selling vs low-selling)
3. Assess sales trends by product (top-selling vs low-selling)
4. Determine the number of orders generated by month
5. Compare monthly revenue and monthly orders
6. Determine monthly average order value 
7. Determine most valuable customer(s)
8. Determine product popularity (by country)
9. Provide an high-level view of sales performance

```{r}
# Install packages if they are not already installed
# install.packages(c("readxl", "dplyr", "ggplot2", "lubridate", #"stringr"))

# Load packages
library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)
library(stringr)
library(readxl)
```

```{r}
# Read dataset
df <- read_excel("C:/Users/dande/Downloads/Online Retail.xlsx")
```

```{r}
# Display top rows
head(df)                  
```
*Explanation:* The first six rows of the dataset.

```{r}
# Display bottom rows
tail(df)
```
*Explanation:* The last six rows of the dataset.

```{r}
# Examine dataset structure
str(df)
```
*Explanation*
The columns in the dataset & their corresponding datatypes are: 
- InvoiceNo: character
- StockCode: character
- Description: character
- Quantity: numeric
- InvoiceDate: datetime
- UnitPrice: numeric
- CustomerID: numeric
- Country: character

```{r}
# Display no. of rows and columns
dim(df)
```
*Explanation:* The dataset has 541909 rows and 8 columns.

```{r}
# Display descriptive stats
summary(df)
```
*Explanation:* For columns of 'character' datatype, statistics of length, number of unique values, number of blank values, minimum character length and maximum character length are shown. Conversely, for columns of 'numeric' and 'datetime' datatype, the minimum, 1st quartile (25% mark), median, mean, 3rd quartile (75% mark) and maximum are shown. 

```{r}
# Count duplicated rows
sum(duplicated(df))
```
*Explanation:* 5268 duplicate rows were identified in the dataset. If kept in the analysis, these will produce inflated sales performance results.

```{r}
# Count missing values
colSums(is.na(df))
```
*Explanation:* 'Description' and 'CustomerID' columns have 1454 and 135080 missing (NA) values, respectively.

```{r}
# Count transactions with negative quantities
sum(df$Quantity < 0)

# Display some negative-quantity transactions
head(
  df[df$Quantity < 0, ],
  10
)
```
*Explanation:* Rows with negative quantities amounted to 10624. If kept in the analysis, these will provide unreliable results of the retail store's sales performance.

```{r}
# Mass Data Cleaning
df_clean <- df %>%
  
  # Remove duplicate rows
  distinct() %>%
  
  # Remove rows without a product description
  filter(!is.na(Description)) %>%
  
  # Keep only positive quantities
  filter(Quantity >= 1) %>%
  
  # Keep only positive prices
  filter(UnitPrice >= 1) %>%
  
  # Remove non-product/service descriptions
  filter(
    !Description %in% c(
      "?",
      "Adjust bad debt",
      "amazon",
      "AMAZON FEE",
      "Bank Charges",
      "check",
      "CARRIAGE",
      "CRUK Commission",
      "damages",
      "DOTCOM POSTAGE",
      "Discount",
      "Manual",
      "Next Day Carriage",
      "POSTAGE",
      "SAMPLES",
      "samples"
    )
  ) %>%
  
  # Remove cancelled invoices beginning with C
  filter(!str_starts(InvoiceNo, "C")) %>%
  
  # Remove adjustments beginning with A
  filter(!str_starts(InvoiceNo, "A"))
```
*Explanation:* Data removed were:
1. Duplicate rows
2. Rows without a product description
3. Rows with -/0 quantities
4. Rows with -/0 unit prices
5. Rows unrelated to products
6. Rows for cancelled invoices
7. Rows for adjustments

```{r}
# View updated dataset structure
dim(df_clean)
```
*Explanation:* After cleaning, 413015 rows and 8 columns are remaining.

```{r}
# Boxplot of Quantities
ggplot(df_clean, aes(x = Quantity)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Quantity",
    x = "Quantity"
  ) +
  theme_minimal()
```
*Explanation:* The boxplot of 'Quantity' column showed 2 quantities deviated greatly from normal. These if kept in the analysis will also provide unreliable results.

```{r}
# Boxplot of Unit Prices
ggplot(df_clean, aes(x = UnitPrice)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Unit Price",
    x = "Unit Price"
  ) +
  theme_minimal()
```
*Explanation:* The boxplot of 'Unit Price' column proved that 2 unit prices deviated greatly from normal. These if kept in the analysis will also provide unreliable results.

```{r}
# Find Quartiles of Quantity
Q1_quantity <- quantile(df_clean$Quantity, 0.25, na.rm = TRUE)
Q3_quantity <- quantile(df_clean$Quantity, 0.75, na.rm = TRUE)

IQR_quantity <- Q3_quantity - Q1_quantity

IQR_quantity
```
*Explanation:* The interquartile range for 'Quantity' of 7 proves that there is a small spread of the middle 50% of the data.

```{r}
# Find Quartiles of Unit Price
Q1_price <- quantile(df_clean$UnitPrice, 0.25, na.rm = TRUE)
Q3_price <- quantile(df_clean$UnitPrice, 0.75, na.rm = TRUE)

IQR_price <- Q3_price - Q1_price

IQR_price
```
*Explanation:* The interquartile range for 'Unit Price' of 3.3 proves that there is a very small spread of the middle 50% of the data.

```{r}
# Calculate lower and upper limits for Quantity
quantity_lower <- Q1_quantity - 1.5 * IQR_quantity
quantity_upper <- Q3_quantity + 1.5 * IQR_quantity

# Calculate lower and upper limits for UnitPrice
price_lower <- Q1_price - 1.5 * IQR_price
price_upper <- Q3_price + 1.5 * IQR_price

# Remove observations outside the limits
df_clean <- df_clean %>%
  filter(
    Quantity >= quantity_lower,
    Quantity <= quantity_upper,
    UnitPrice >= price_lower,
    UnitPrice <= price_upper
  )
```
*Explanation:* Rows outside of the interquartile boundaries for 'Quantity' and 'Unit Price' were excluded to avoid the extreme values from influencing the analysis.

```{r}
# Display updated dataset structure
dim(df_clean)
```
*Explanation:* After dropping the rows with extreme values, 356207 rows and 8 columns are remaining.

```{r}
# Count unique countries
n_distinct(df_clean$Country)
```
*Explanation:* 38 different countries were identified in the dataset.

```{r}
# Display all unique products
unique_products <- df_clean %>%
  select(StockCode, Description) %>%
  distinct() %>%
  arrange(Description)

# Display the unique products
unique_products
```

```{r}
# Calculate revenue for each transaction
df_clean$Sales <- df_clean$Quantity * df_clean$UnitPrice

# Display the first few records
head(df_clean[, c(
  "InvoiceNo",
  "Description",
  "Quantity",
  "UnitPrice",
  "Sales"
)])
```
*Explanation:* The revenue, 'Quantity' x 'UnitPrice', was calculated to provide the retail store's management a sound basis for determining sales performance besides number of products sold.

```{r}
# Convert InvoiceDate to Date format
df_clean$Date <- as.Date(
  df_clean$InvoiceDate
)

# Display the first few dates
head(df_clean$Date)
```
*Explanation:* The 'InvoiceDate' column has both date and time values. Only the date aspect is needed for further analysis to see how sales changes over time.

```{r}
# Find the earliest transaction date
min(df_clean$Date)

# Find the latest transaction date
max(df_clean$Date)
```
*Explanation:* The earliest and latest dates in the dataset are examined to become familiar with the date range of the analysis and for better understanding of sales growth and/or decline patterns.

```{r}
# Calculate daily sales
daily_sales <- df_clean %>%
  group_by(Date) %>%
  summarise(
    Total_Sales = sum(Sales, na.rm = TRUE)
  )

# Display the first records
head(daily_sales)
```

*Explanation:* Through assessing daily sales, it be an indication to the retail store's owners of the typical day(s) vs non-typical days for purchases. 

```{r}
# Display daily sales
ggplot(
  daily_sales,
  aes(
    x = Date,
    y = Total_Sales
  )
) +
  geom_line() +
  labs(
    title = "Daily Sales Revenue Trend",
    x = "Date",
    y = "Sales Revenue (£)"
  ) +
  theme_minimal()
```
*Explanation:* The daily sales revenue trend over 2010-2011 show primarily sharp rises and falls in sales. There is a staggered fall in sales in late 2010 and a gradual yet staggered increase in sales in late 2011 (roughly Nov to Dec). For the retail store, these rises and falls could be due to rapidly changing customer demands, attractive and/or unattractive sales promotions, inventory shortages or operational disruptions (e.g. website/payment system downtime, delivery delays).

```{r}
# Calculate monthly sales
df_clean$YearMonth <- format(
  df_clean$Date,
  "%Y-%m"
)

monthly_sales <- df_clean %>%
  group_by(YearMonth) %>%
  summarise(
    Total_Sales = sum(
      Sales,
      na.rm = TRUE
    )
  )

head(monthly_sales)
```
*Explanation:* Through assessing monthly sales, the retail store's owners can use this data for more strategic planning (eg. increasing inventory given repeated demand) and to be informed of sales influenced by seasons (eg. shopping for national holidays like Christmas).

```{r}
# Plot monthly sales
ggplot(
  monthly_sales,
  aes(
    x = YearMonth,
    y = Total_Sales,
    group = 1
  )
) +
  geom_line() +
  geom_point() +
  labs(
    title = "Monthly Sales Revenue Trend",
    x = "Month",
    y = "Sales Revenue (£)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```
*Explanation:* The monthly sales revenue trend for the retail store shows that it had lower sales in Dec 2010 to Aug 2011. This pattern could be attributed to weak customer demand and/or promotion by the retail store. Its sales increased after Aug 2011 to Nov 2011 then declined dramatically again in Dec 2011. This is an expected demand pattern nearing holiday seasons except for the decline in Dec 2011 which needs more context to explain. 

```{r}
# Identify the best sales months
top_months <- monthly_sales[
  order(
    monthly_sales$Total_Sales,
    decreasing = TRUE
  ),
]

head(top_months, 10)
```
*Explanation:* The best sales months were primarily in the 3rd and 4th quarters of the years. Though, not certain, this could be due to greater customer demand (local and international) given additional promotions, new product launches or changes to customer buying patterns.

```{r}
# Identify the weakest sales months
bottom_months <- monthly_sales[
  order(
    monthly_sales$Total_Sales,
    decreasing = FALSE
  ),
]

head(bottom_months, 10)
```
*Explanation:* The weakest sales months were mainly in the 2nd and 3rd quarters of the years. This finding could be further compared to inventory levels, product options, promotion efforts, operational problems (if any) and customer buying patterns at that time for greater context. 

```{r}
# Calculate sales by year
df_clean$Year <- format(
  df_clean$Date,
  "%Y"
)

yearly_sales <- df_clean %>%
  group_by(Year) %>%
  summarise(
    Total_Sales = sum(
      Sales,
      na.rm = TRUE
    )
  )

yearly_sales
```
*Explanation:* This yearly sales data provides the retail store's owners with a high level view of revenue. However, as 2010 only has one month represented in the dataset, this imbalance gives favour to higher sales in 2011. Consequently, the results are unsuitable for determining true demand for either years.

```{r}
# Plot yearly sales
ggplot(
  yearly_sales,
  aes(
    x = Year,
    y = Total_Sales,
    group = 1
  )
) +
  geom_line() +
  geom_point() +
  labs(
    title = "Annual Sales Revenue Trend",
    x = "Year",
    y = "Sales Revenue (£)"
  ) +
  theme_minimal()
```
*Explanation:* The data shows an increase in yearly sales from 2010 to 2011 for the retail store; almost 12 times that of 2010. Yet, this is fostered by a dataset imbalance.

```{r}
# Analyze sales by day of the week
df_clean$DayOfWeek <- weekdays(
  df_clean$Date
)

day_sales <- df_clean %>%
  group_by(DayOfWeek) %>%
  summarise(
    Total_Sales = sum(
      Sales,
      na.rm = TRUE
    )
  )

day_sales
```
*Explanation:* An analysis of the sales per day of the week helps the retail store's owners to identify the weekday(s) that see the most versus least revenue. The store's owners may then dynamically adjust staffing and efficiency to fulfill orders, timing of promotions and advertising and inventory reorder schedules.

```{r}
# Put weekdays in the correct order
day_sales$DayOfWeek <- factor(
  day_sales$DayOfWeek,
  levels = c(
    "Monday",
    "Tuesday",
    "Wednesday",
    "Thursday",
    "Friday",
    "Saturday",
    "Sunday"
  )
)

day_sales <- day_sales[
  order(day_sales$DayOfWeek),
]
```

```{r}
# Plot sales by day of week
ggplot(
  day_sales,
  aes(
    x = DayOfWeek,
    y = Total_Sales
  )
) +
  geom_bar(
    stat = "identity"
  ) +
  labs(
    title = "Sales Revenue by Day of Week",
    x = "Day of Week",
    y = "Sales Revenue (£)"
  ) +
  theme_minimal()
```
*Explanation:* Tuesdays and Thursdays reflect similar revenue and the highest revenues. Mondays, Wednesdays and Fridays also had similar but lower revenue with Sunday having the lowest revenue. The retail store's owners may apply any or some of the strategies as previously suggested to capitalize on the "high revenue" days or boost sales on "low revenue" days.

```{r}
# Analyze monthly seasonality
df_clean$Month <- format(
  df_clean$Date,
  "%m"
)

monthly_seasonality <- df_clean %>%
  group_by(Month) %>%
  summarise(
    Total_Sales = sum(
      Sales,
      na.rm = TRUE
    )
  )

monthly_seasonality
```

*Explanation:* Monthly seasonality helps the retail store's owners identify which month(s) overall had the best sales, specifically seasonality patterns. Therefore, the owners could implement strategies including acquiring adequate stock for the season, increase advertising/marketing efforts and optimizing order fulfillment operations ahead of these peak seasons.

```{r}
# Order the months correctly
monthly_seasonality$Month <- factor(
  monthly_seasonality$Month,
  levels = c(
    "01", "02", "03", "04",
    "05", "06", "07", "08",
    "09", "10", "11", "12"
  )
)

monthly_seasonality <- monthly_seasonality[
  order(monthly_seasonality$Month),
]
```

```{r}
# Plot monthly seasonality
ggplot(
  monthly_seasonality,
  aes(
    x = Month,
    y = Total_Sales
  )
) +
  geom_bar(
    stat = "identity"
  ) +
  labs(
    title = "Sales Revenue by Calendar Month",
    x = "Month",
    y = "Total Sales Revenue (£)"
  ) +
  theme_minimal()
```
*Explanation:* Considering sales growth from Sept to Dec, with Nov having the highest sales, this is an indication of seasonality. These levels could be attributed to increased customer buying for holidays like Thanksgiving and Christmas celebrated during that time.

```{r}
# Analyze international sales
country_sales <- df_clean %>%
  group_by(Country) %>%
  summarise(
    Total_Sales = sum(
      Sales,
      na.rm = TRUE
    )
  ) %>%
  arrange(
    desc(Total_Sales)
  )

head(country_sales, 10)
```
*Explanation:* Most of the online retail store's sales are generated from its base country, the United Kingdom. This may be in its owners' best favour as their marketing and shipping efforts can be minimal. However, this also presents a grand opportunity for greater international, country-specific marketing and improving shipping operations, therefore catering to a wider customer base.

```{r}
# Plot the top 10 countries
top_countries <- head(
  country_sales,
  10
)

ggplot(
  top_countries,
  aes(
    x = reorder(
      Country,
      Total_Sales
    ),
    y = Total_Sales
  )
) +
  geom_bar(
    stat = "identity"
  ) +
  coord_flip() +
  labs(
    title = "Top 10 Countries by Sales Revenue",
    x = "Country",
    y = "Sales Revenue (£)"
  ) +
  theme_minimal()
```
*Explanation:* The United Kingdom's revenue is the millions while its runner ups' are in the hundreds of thousands. These countries (Netherlands, EIRE, Germany, France) are however good targets for further marketing efforts by the retail store to drive their revenue close to or matching that of the United Kingdom.

```{r}
# Find the highest selling products
product_sales <- df_clean %>%
  group_by(
    StockCode,
    Description
  ) %>%
  summarise(
    Total_Sales = sum(
      Sales,
      na.rm = TRUE
    )
  ) %>%
  arrange(
    desc(Total_Sales)
  )

head(product_sales, 10)
```
*Explanation:* An analysis of the overall revenue per product can aid the retail store's owners in identifying the popular products, focusing inventory restock efforts, further promoting the "high sales" products and investigating the "low sales" products. 

```{r}
# Plot the top 10 products
top_products <- head(
  product_sales,
  10
)

ggplot(
  top_products,
  aes(
    x = reorder(
      Description,
      Total_Sales
    ),
    y = Total_Sales
  )
) +
  geom_bar(
    stat = "identity"
  ) +
  coord_flip() +
  labs(
    title = "Top 10 Products by Sales Revenue",
    x = "Product",
    y = "Sales Revenue (£)"
  ) +
  theme_minimal()
```
*Explanations:* The retail store owner's sales efforts across international markets could be centred around these 10 products to capitalize on their high revenue potential.

```{r}
# Calculate total sales for each product within each country
country_product_sales <- df_clean %>%
  group_by(Country, StockCode, Description) %>%
  summarise(
    Total_Sales = sum(Sales, na.rm = TRUE),
    .groups = "drop"
  )

# View the results
head(country_product_sales)
```
*Explanation:* Through examining the product sales per country, the retail store's owners can better understand the demand distribution. Therefore, they can optimize inventory and shipping operations to ensure these products are readily available and delivered on time.  

```{r}
# Find the highest-selling product in each country
highest_product_by_country <- country_product_sales %>%
  group_by(Country) %>%
  slice_max(
    order_by = Total_Sales,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup()

# Display the results
highest_product_by_country
```
*Explanation:* The highest-selling product per country provides the retail store's owners with insight into which product contributes most to its revenue. Its owners may then employ additional advertising strategies (eg. social media/tv ads) to drive more sales of these products.

```{r}
# Calculate quantity sold for each product within each country
country_product_quantity <- df_clean %>%
  group_by(Country, StockCode, Description) %>%
  summarise(
    Total_Quantity = sum(Quantity, na.rm = TRUE),
    .groups = "drop"
  )

# Find the product with the highest quantity sold in each country
highest_quantity_by_country <- country_product_quantity %>%
  group_by(Country) %>%
  slice_max(
    order_by = Total_Quantity,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup()

highest_quantity_by_country
```
*Explanation:* The highest-selling product per country and product with highest quantity sold per country differ. Yet, the retail store's owners can still capitalize on maintaining adequate inventory levels as these products still contribute to revenue though to a smaller degree.  

```{r}
# Calculate the number of orders by month
monthly_orders <- df_clean %>%
  group_by(YearMonth) %>%
  summarise(
    Number_of_Orders = length(
      unique(InvoiceNo)
    )
  )

head(monthly_orders)
```
*Explanation:* The number of orders per month aids the retail store's owners in differentiating between more customers or more orders versus larger orders from existing customers.

```{r}
# Compare monthly revenue with monthly orders
monthly_summary <- merge(
  monthly_sales,
  monthly_orders,
  by = "YearMonth"
)

head(monthly_summary)
```
*Explanation:* The retail store's owners can get a better understanding of monthly sales performance through a side-by-side view of sales and number of orders. From a glance at this data, it appears a few orders generate much revenue per month.

```{r}
# Calculate average order value
monthly_summary$Average_Order_Value <-
  monthly_summary$Total_Sales /
  monthly_summary$Number_of_Orders

head(monthly_summary)
```
*Explanation:* The data shows that based on the first six results, that each order generated on average £300 (rounded). In other words, this is the average spend per customer on an order. As the number of orders and average order value fluctuate, revenue is driven by both more orders and greater customer spend and not just one factor.

```{r}
# Plot average order value
ggplot(
  monthly_summary,
  aes(
    x = YearMonth,
    y = Average_Order_Value,
    group = 1
  )
) +
  geom_line() +
  geom_point() +
  labs(
    title = "Monthly Average Order Value",
    x = "Month",
    y = "Average Order Value (£)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```
*Explanation:* The data shows that customers spent varying amounts on orders over 2010-2011. The greatest spend were in Jan 2011 and Oct 2011, with a drastic decrease from Feb 2011 to Jul 2011. The low points were prime opportunities for the store's owners to implement product bundling, cross-selling/upselling, discounts and other promotions to improve sales.

```{r}
# Calculate customer-level sales measures
customer_value <- df_clean %>%
  filter(!is.na(CustomerID)) %>%
  group_by(CustomerID) %>%
  summarise(
    Total_Sales = sum(Sales, na.rm = TRUE),
    Number_of_Orders = n_distinct(InvoiceNo),
    Total_Quantity = sum(Quantity, na.rm = TRUE),
    Average_Order_Value = Total_Sales / Number_of_Orders,
    .groups = "drop"
  )

# Find the customer with the highest total sales
most_valuable_customer <- customer_value %>%
  slice_max(
    order_by = Total_Sales,
    n = 1,
    with_ties = FALSE
  )

most_valuable_customer
```
*Explanation:* Customer with ID# 14911 was identified as the "most valuable customer" during the period and therefore could be extended special offers like free shipping once per month or coupons to encourage their continued patronage to the retail store.

```{r}
# Find the top 10 customers by total sales
top_10_customers <- customer_value %>%
  arrange(desc(Total_Sales)) %>%
  head(10)

# Display the results
top_10_customers
```
*Explanation:* These customers constitute the retail store's "high-revenue customer group". Likewise, they could be extended special offers to maintain their loyalty.

```{r}
# Plot top 10 customers by total sales
ggplot(
  top_10_customers,
  aes(
    x = reorder(
      as.factor(CustomerID),
      Total_Sales
    ),
    y = Total_Sales
  )
) +
  geom_bar(
    stat = "identity"
  ) +
  coord_flip() +
  labs(
    title = "Top 10 Customers by Total Sales",
    x = "Customer ID",
    y = "Total Sales (£)"
  ) +
  theme_minimal()
```
*Explanation:* The high revenue experienced by the online retail store can be credited to Customer with ID# 14911, as this customer's sales far exceed the others in the high-revenue customer group. Data on this customer's purchasing frequency would further substantiate this finding.

```{r}
# Useful final management summary
total_revenue <- sum(
  df_clean$Sales,
  na.rm = TRUE
)

total_orders <- length(
  unique(df_clean$InvoiceNo)
)

total_products_sold <- sum(
  df_clean$Quantity,
  na.rm = TRUE
)

average_order_value <-
  total_revenue / total_orders

cat(
  "Total Revenue: £",
  round(total_revenue, 2),
  "\n"
)

cat(
  "Total Orders:",
  total_orders,
  "\n"
)

cat(
  "Total Products Sold:",
  total_products_sold,
  "\n"
)

cat(
  "Average Order Value: £",
  round(average_order_value, 2),
  "\n"
)
```
*Explanation:* Over the course of 2010-2011, the online retail store earned £4,341,542 in total revenue from 17615 orders of 1,580,985 products valued at £246.47 on average per order.
